---
title: "Process_cover_data"
author: "Nicole"
date: "3/11/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = FALSE}
rm(list = ls())
```

Use this`getwd()` to check working directory. Will most likely be whatever folder your Rmd file is in. If you need to change it for some reason use `setwd()` 
```{r, echo = FALSE, include=FALSE}
getwd()
#setwd()
```

# Downloading/loading packages

Use `install.packages` to download and install a new R package. You only need to do this once, then you can just use `library()` to load it and be able to call functions within it. and load the package libraries.  

**TIP: It's generally considered best practice to keep the commands to call all the packages you need to use at the top of the document**  

`tidyr`, `dplyr`, and `readr` are probably the most commonly used packages for data wrangling/cleaning.`ggplot2` is very common for creating publication quality figures.  

**Note: The order which you install packages is important. Functions with the same names will be masked by packages loaded in after the previous.**  
Ex: The `select()` function from `MASS` will be masked by `select()` from `dplyar` 
```{r, message=FALSE, warning=FALSE}
library("leaflet")
library("MASS") #used for post-hoc test (lda)
library("tidyr")
library("dplyr")
library("readr")
library("ggplot2")
library("ggpubr")
```

# Data Wrangling

### Load in data and merge datasets:

Load in each years .csv files containing the data
```{r}
test_data <- read.csv("../Data/Test_data.csv", header = TRUE)
```

Look at the data to see what it contains  
*Make sure the column names in each data set match so they can be merged together*
```{r, message=FALSE, results='hide'}
head(test_data) #opens first 6 rows
tail(test_data) #opens last 6 rows
summary(test_data)
str(test_data) #tells us what data types (numbers, factors, etc) are in the data frame
```


### Transform from wide to long format
```{r}
test_long <- test_data %>%
  filter(date != "drop") %>%
  pivot_longer(cols = -c(date, year, region, site, treatment, block, species), names_to = "quad_num", values_to = "cover") 
```

### Remove any data that you will not need for the analysis

In this case we can remove unknowns, vole activity, tussock number...ect.
```{r}
#Returns a list of unique species names. Use this to see what you want to eliminate. Rerun after you remove data to make sure it worked. 
unique(test_long$species) 

#Remove data you don't need/want
test_long <- test_long %>% filter(species != "Unk.N2" & species != "vole hole (#)" & species != "chopped vole litter " & species != "tussock #" & species != "sampled " & species != "latrienes (%)" & species != "vole trail (%)" & species != "Severed vole litter" & species != "trampled ")
```

### Fix naming convention errors

Rename "rel.cov" column to "rel_cov"
```{r}
#relcov_08_21 <- relcov_08_21 %>%
  #rename(rel_cov = rel.cov)
```

Check unique vales in each column to make sure that there are not naming errors
```{r, results='hide'}
unique(relcov_08_21$species)
```

If there are mistakes then rename them using the code below and recheck unique values again to make sure the recode worked. *make sure you didn't loose any observations*
```{r}

#fix naming convention errors 
relcov_08_21$species <- relcov_08_21$species %>% 
  recode("bare " = "bare", "Litter" = "litter", "Moss" = "moss", "Lichen " = "lichen","Lichen" = "lichen", "Fr boil" = "frost boil", "St. D. Sal pul" = "St. D. Sal", "Ev litter" = "Eri vag litter", "Grass ex." = "grass")

#simplify things identified to species level  that shouldn't be 
  #St.D.
relcov_08_21$species <- relcov_08_21$species %>% 
recode("St. D. Led Pal" = "St. D", "Dead Evrg" = "St. D", "Dead Bet" = "St. D", "St. D. Bet" = "St. D", "St. D. Sal" = "St. D", "St. D. Bet." = "St. D", "St. D. Evrg" = "St. D")
  #litter
relcov_08_21$species <- recode(relcov_08_21$species, "Eri vag litter" = "litter")
```

Sum functional cover across species within quadrats. This likely wont change the relative cover for most species unless they were listed more than once in a quadrat but it's good to do to insure accuracy after fixing naming conventions.
```{r}
relcov_08_21 <- (relcov_08_21) %>% group_by(year, region, site, block, plot, quad, species) %>% summarise_at(vars(rel_cov), list(rel_cov = sum ), na.rm = TRUE)
```



### Relativize cover

1. First we need to sum cover values across all species within a quadrat
```{r}
test_quadsum <- (test_long) %>% 
  group_by(date, region, treatment, block, quad_num) %>%
  summarise(sum_quad = sum(cover)) %>%
  ungroup()
```
2. Then we join the new table (that has the sum of all the species cover for a quadrat), with original table
```{r}
test_join <- left_join(test_long, test_quadsum, by= c("date", "region", "site", "treatment", "block", "quad_num"))
```
3. Then we divide each cover value for a species in a quadrat by the sum of all cover values in that quadrat, and then drop the old 
```{r}
test_clean <- test_join %>%
  mutate(relcov = cover/sum_quad) %>%
  select(-sum_quad, -cover)
```

































